Efficient Compilation of High Level Python Numerical Programs with Pythran

### 2014-08-25 doc/papers/cgo2014/paper.tex
Add section and ref
Typo corrections

    As more scientists use these scripting language, there is an on-going
    research effort to compile them into efficient native code. However the
    computations kernels often differ from classical ones as these languages
    favors an high-level coding style, where loops are often replaced by calls
    to the relevant toolbox/module/library/package. As a consequence, classical
    loop optimization techniques like loop fusion, tiling and related
    vectorization or parallelization techniques are not directly applicable.

    This papers focuses on the case of the Python language and especially the
    Numpy package that provides core array data structure and basic linear
    algebra routines. It first conducts a case study based on Numpy user
    questions from the stackoverflow question and answer website, then studies
    existing compilers for numerical Python. Optimization opportunities,
    including parallelization and vectorization, and their implementation in
    the Pythran compiler are presented and illustrated through several
    benchmarks, showing very interesting speedup over the standard Python
    interpreter. Speedup of factor greater than 10 are achieved despite the
    fact the considered benchmarks mostly call Numpy routines that run C code.

@@

One of the first steps required when designing an optimizing compiler is to
gather enough test cases, or benchmarks, in order to drive the optimization
process toward realistic examples. To achieve this goal, we have gathered a few
exemple from existing compiler sources, but these may have been modified to
overcome some compiler limitations. A less biased source of benchmarks should
be independant from the compiler community and focused on matching user needs.
To achieve that goal, we collected synthetic scientifc kernels from the
stackoverflow website.

@@

For instance, question \url{http://stackoverflow.com/questions/7741878}
proposes to use an explicit loop to iteratively compute the $L^2$ norm of each
row of a matrix, as illustrated in Listing~\ref{lst:l2norm-loopy}, and one of
the proposed answer is to use the more compact version presented in
Listing~\ref{lst:l2norm}. This answer no longer uses any explicit loops and
roughly achieves a $times 4$ speedup over the loop version.

@@

Our guess is that new Numpy user are more likely to write loops and code
similar to Listing~\ref{lst:l2norm-loopy}, but as their knowledge of the numpy
API and of the good programming practice grows, they get to write code similar
to Listing~\ref{lst:l2norm}. A direct consequence is that a compiler should
focus on optimizing high-level constructs rather loops, both to encourage good
practice with respect to Numpy code, and because that's the kind of code Numpy
user ultimately write.

@@

@@

\cite{expression_templates}

>>>>

    As more scientists use these scripting languages, there is an on-going
    research effort to compile them into efficient native code. However the
    computation kernels often differ from the classical ones as these languages
    favor a high-level coding style, where loops are often replaced by calls
    to the relevant toolbox/module/library/package. As a consequence, traditional
    loop optimization techniques like loop fusion, tiling and related
    vectorization or parallelization techniques are not directly applicable.

    This papers focuses on the case of the Python language and especially the
    Numpy package that provides core array data structure and basic linear
    algebra routines. It first conducts a case study based on Numpy use
    cases from the stackoverflow question and answer website, then studies
    existing compilers for numerical Python. Optimization opportunities,
    including parallelization and vectorization, and their implementation in
    the Pythran compiler are presented and illustrated through several
    benchmarks, showing very interesting speedups over the standard Python
    interpreter. Speedup greater than a factor of 10 are achieved despite the
    fact that the considered benchmarks mostly call Numpy routines that run C code.

@@

One of the first steps required when designing an optimizing compiler is to
gather enough test cases, or benchmarks, in order to drive the optimization
process toward realistic examples. To achieve this goal, we have gathered a few
example from existing compilers' testsuite, but these may have been modified to
overcome some compiler limitations. A less biased source of benchmarks should
be independant from the compiler community and focused on matching user needs.
To achieve that goal, we collected synthetic scientifc kernels from the
stackoverflow website.

@@

For instance, question \href{http://stackoverflow.com/questions/7741878}{7741878}
proposes to use an explicit loop to iteratively compute the $L^2$ norm of each
row of a matrix, as illustrated in Listing~\ref{lst:l2norm-loopy}, and one of
the proposed answer is to use the more compact version presented in
Listing~\ref{lst:l2norm}. This answer no longer uses any explicit loops and
roughly achieves a $times 4$ speedup over the loop version.

@@

Our guess is that beginner Numpy users are more likely to make use of raw loops
similar to Listing~\ref{lst:l2norm-loopy}, but as their knowledge of the numpy
API and of the good programming practice grows, they get to write code similar
to Listing~\ref{lst:l2norm}. As a direct consequence, a compiler should
focus on optimizing high-level constructs rather than raw loops. Both to encourage good
practice with respect to Numpy code, and because this is the kind of code Numpy
users ultimately write.

@@

\subsection{Typing}

@@

\cite{expression_templates, et2012}
